Validation of Visual Statistical Inference, Applied to Linear Models

### 2012-06-27 paper_jasa.tex
small comment

\paragraph{Subject-specific abilities}
Calculating the power  is intricately related to the probability to identify the data plot from the lineup.

@@

Let $X_{\ell i} \sim Binom_{1, p_{\ell i}}$ be the random variable, with $X_{ \ell i}= 1$, if  subject $i$ correctly identifies the data plot from lineup $\ell$, and 0 otherwise.
We can model $p_{ \ell i}$ with a logistic regression with subject specific random effects as:
\begin{equation} \label{mixed}
g( p_{\ell i} )= W_{\ell i} \delta +  Z_{\ell i}  \tau_{\ell i},  
\end{equation}

@@

\begin{tabular}{lp{5.5in}}
$W$  & is a design matrix of covariates corresponding to specifics of lineup $\ell$ and subject $i$. Covariates could include  demographic information of individuals, such as age, gender, education level etc. as well lineup-specific elements, e.g. difficulty level. \\
$\delta$ & is a vector of parameters for fixed effects.\\
$Z_{\ell i}$ & $1 \le i \le K$, $1 \le \ell \le L$  is a design matrix corresponding to random effects specific to individual $i$ and lineup $\ell$.  \\
$\tau$ & is a vector of independent real-valued random variables $\tau_{\ell i}$,  $\tau  \sim  MVN(0,\sigma_\tau I_{KL \times KL})$. $\tau$ will usually include a component incorporating an individual's ability or skill to evaluate lineups.\\
\end{tabular}

%\red{This model is  described in more detail in equation \ref{mixed}.}

@@

\noindent
\paragraph{Lineup size $m$}
One of the big differences between visual inference and the conventional testing situation is the use of a finite number $m-1$ of representatives of the null distribution as comparison against the test statistic. The choice of $m$ has an obvious impact on the test, and we will discuss this in the following paragraph.

@@

%Thus the density of $p_D$ is Uniform(0,1)  under $H_0$. As noted by \cite{Ruppert:2007}, under $H_1$ the density of $p_D$ $$f_{p_D}(p_D; \delta)= \frac{f_{|t|}(F_{|t|}^{-1}(1-p_D);\delta)}{f_{|t|}(F_{|t|}^{-1}(1-p_D))}$$ derived from equation \ref{dist_p}  is a right skewed distribution.

@@

Figure \ref{fig:pval_power} gives an overview of the probability of picking the actual data plot for lineups of different size: as $m$ increases we have an increased probability to observe a more highly structured null plot by chance. We see that for a $p$-value $p_D$ of about 0.15 for the data plot the signal in the plot is so weak, that we cannot distinguish the actual data plot from null plots in a lineup of size $m=20$.  This pattern is confirmed by our findings described in Section~\ref{sec:simulation}.
%also seen in our experiments as shown in figure \ref{fig:pval_pcorrect}.

@@

Note that the \hh{left}  hand side in Equation \ref{eqn:power_estimate} is independent of the parameter $\theta$, but just based on the lineup evaluation by independent observers. While this allows us to compute this value independently of the parametric setting, it also means that observers might pick out the data plot for a reason not related to the value of $\theta$.

@@

%From model \ref{mixed} we obtain %$\pi=Pr(Y=1|X_1, X_2, ..., X_p)$, 
%the power of the underlying testing procedure as a population average %by ignoring the random effect part. 
%%Thus we have the overall probability of success ($\pi$ or power) 
%for specified sample size ($n$) and  variance ($\sigma$)  as 
%
%\begin{equation}\label{eqn:power} 
%Power(\beta) = \pi=Pr(Y=1|\beta, n, \sigma) 
%\end{equation}
%
%The main covariate factors that determine the power of the test are sample size ($n$), $\sigma$ and the parameter $\beta$. combining these we obtain one single covariate effect denoted by $E$ and defined as $E = \beta/ \sigma*\sqrt{n}$. Using equation \ref{mixed}, we then have the power as
%
%\begin{equation}\label{eqn:effect} 
%Power(E) = \pi=Pr(Y=1| E) 
%\end{equation}

@@

% Di: This is specific for the regression model
%Consider the hypothesis test $H_0: \beta_k=0$ against $H_1: \beta_k \ne 0$ in model \ref{multi}.
%%Now consider estimating the power of the visual test. 
%%Suppose $F_{|t|}(.)$ is the distribution function of an absolute value of $t$. For the regression slope parameter $\beta$ suppose the associated test statistic in classical test be $t_{obs}$ with $p$-value $p_D$. By definition we have 
%%
%%$$p_D=Pr(|t| \ge t_{obs}| H_0)=1-F_{|t|}(t_{obs}) \Rightarrow |t_{obs}|=F_{|t|}^{-1}(1-p_D)$$
%%
%%\red{HH: the following paragraph should probably move up, but I'm not sure how to change it appropriately. Mahbub, could you take care of that?}
%%Now suppose $F_{|t|}(.;\delta)$ denotes the distribution function of an absolute value of $t$ distribution with non-centrality parameter $\delta$. Thus the distribution function of $p-value$ be  
%%\begin{eqnarray}\label{dist_p}
%%F{p_D}(p) &=& Pr(p_D \le p)=1-Pr(1-p_D \le 1-p) \nonumber \\
%%  &=& 1-Pr(F_{|t|}^{-1}(1-p_D) \le F_{|t|}^{-1}(1-p) ) \nonumber \\
%%  &=& 1-Pr(|t_{obs}| \le F_{|t|}^{-1}(1-p)) \nonumber \\
%%  &=&\left\{ \begin{array}{ll}
%%          1-F_{|t|}( F_{|t|}^{-1}(1-p))=p &\mbox{ ; under $H_0$} \\
%%          1-F_{|t|}( F_{|t|}^{-1}(1-p); \delta) &\mbox{ ; under $H_1$} 
%%       \end{array} \right.     
%%\end{eqnarray}
%%
%%Thus the density of $p_D$ is Uniform(0,1)  under $H_0$. As noted by \cite{Ruppert:2007}, under $H_1$ the density $$f_{p_D}(p_D; \delta)= \frac{f_{|t|}(F_{|t|}^{-1}(1-p_D);\delta)}{f_{|t|}(F_{|t|}^{-1}(1-p_D))}$$ derived from equation \ref{dist_p}  is a right skewed distribution.  
%In a lineup plot we simulate $m-1$ residual data sets from null model where each of these $m-1$ null data sets produces a corresponding $p$-value $p_{0,i}$ and $p_{0,i} \sim \text{Uniform}(0,1)$ for $i = 1, ..., m-1$. Suppose  $p_0=\min_i(p_{0,i})$. Thus $p_0 \sim \text{Beta}(1,m-1)$.

@@

\begin{equation}\label{power_exp} 
   Power(\beta)=Pr(p_{D} < p_0)  \quad \text{for}  \quad \beta \ne 0
\end{equation}
Figure \ref{fig:power_expected} shows the power of the conventional test in comparison to the expected power of the visual test obtained from Equation \ref{power_exp}. Notice that the expected power of the visual test is almost as good as the power of conventional test. \green{What do you mean by this? Where does this ``fact'' come from?} In fact, the only loss we experience stems from comparing the test statistic to only a discrete number of  $m-1$ samples from the null distribution.

%Theorem \ref{thm:power} helps to compute the expected power of visual inference.
%\begin{thm}\label{thm:power}
% Equation \ref{power_exp}  yields expected power of visual inference as Power($\beta$) = $E(1-p_D)^{m-1}$ .
%\end{thm}
% 
%{ \em proof}
%
%Since $p_0 \sim \text{Beta}(1,m-1)$,  for $t \in (0,1)$ we have the distribution function of $p_0$ be
%\begin{eqnarray*}
%F{p_0}(t) &=& (m-1) \int_0^t{(1-p_0)^{(m-2)}dp_0} \\
%  &=&  -(m-1) \int_1^{1-t}{x^{(m-2)}dx}\\
%  &=& \left[ x^{m-1}\right]^1_{1-t}\\
%  &=&1-(1-t)^{m-1} \rightarrow 1 \quad \text{as} \quad m \rightarrow \infty
%\end{eqnarray*}
%
%% verified this computation from http://www.wolframalpha.com using command int (m-1)*(1-p)^(m-2) dp, p=0,t
%
%Thus expected power in equation \ref{power_exp} be
%\begin{eqnarray*}
% Pr(p_{B} < p_0) &=& 1- Pr( p_0 \le p_{B} ) \\
%  &=& 1- \int_{0}^{1}{Pr( p_0 \le p_{B} |p_{B} =t) f_{p_{B} }(t)dt } \\
%  &=& 1-  \int_{0}^{1}{F{p_0}(t) f_{p_{B} }(t)dt } \\
%  &=& E(1-p_D)^{m-1}   \\
%%  & \rightarrow &1-  \int_{0}^{1}{ f_{p_{B} }(t)dt }  = F_{p_D}(0) \quad \text{as} \quad m \rightarrow \infty
%\end{eqnarray*}
%
%
\begin{figure}[hbtp]
   \centering
       \scalebox{.6}{\includegraphics{power_expected.pdf}}
       \caption{Expected power of visual test  and the power of conventional test for sample size $n =100$, $\sigma = 12$ and $m$=20. }
       \label{fig:power_expected}
\end{figure}

@@

\begin{figure}[htp]
%\begin{figurehere}
   \centering
       \includegraphics[width=0.95\textwidth]{lineup_continuous.pdf}
       \caption{Lineup plot ($m=20$) using scatter plots for testing $H_0: \beta_k=0$ where covariate $X_k$ is continuous. One of these plots is the plot of the actual data, and the remaining are null plots, produced by simulating data from a null model that assumes $H_0$ is true. Which plot is the most different from the others, in the sense that there is the steepest slope?}
       \label{fig:test_continuous}
\end{figure}

@@

\[
  Y_i = \left\{
  \begin{array}{l l}
    \alpha+\beta X_i + \epsilon_i  & \quad  X_i \sim N(0,1) \quad  i =1,...,n\\
    \lambda+ \eta_i & \quad X_i \sim N(\mu,1/3) \quad  i=n+1, ...,n*c\\
  \end{array} \right.
\]
where $\epsilon_i \stackrel{iid}\sim N(0,\sigma)$, $\eta_i \stackrel{iid}\sim N(0,\sigma/3)$ and $\mu = -1.75$, $c$ is the proportion of contaminated data. For the experiment we consider $n=100$, $c=0.15$ producing actual data with $115$ points. Also $\alpha=0$ and $\lambda=10$, and $\sigma$ is chosen to be approximately 3.5, so that error standard deviation across both groups of the data is $5$. Figure \ref{fig:cont_dat} shows an example of contaminated data, generated by this model for $\beta=5$.  A linear model (Equation \ref{multi} with $p=1$  and intercept $\beta_0=0$) is fit to the contaminated data. This experiment also follows the test in Table \ref{tbl:stat_multiple}, case 2. The actual data plot shows a scatterplot of the residuals vs $X_1$, and the null data plots are scatterplots of null data generated by simulating residuals from $N(0, {\hat{\sigma}}^2$) and plotting against $X_1$.

@@

An example lineup for slope $\beta=0.4$ is shown in Figure \ref{fig:test_contaminated}.  Can you pick which plot is different? %Plot 7 is the actual data plot. 
\begin{figure}[hbt]
%\begin{figurehere}
   \centering
       \scalebox{0.6}{\includegraphics{contaminated_data.pdf}}
       \caption{Scatterplot of a simulated contaminated data set generated such that a simple linear model produce a slope almost negligible, ie, if we fit model (Equation \ref{multi}) with $p=1$  to this data, the resulting estimate of $\beta_1$ would not be statistically significant. The conventional t-test $p$-value for the slope is 0.67 for this data set but visually we can see a convincing evidence of positive slope. 
}
       \label{fig:cont_dat}
\end{figure}

\begin{figure}[htp]
%\begin{figurehere}
   \centering
       \includegraphics[width=0.95\textwidth]{lineup_contaminated.pdf}
       \caption{Lineup plot ($m=20$) using scatter plots for testing $H_0: \beta_k=0$ where covariate $X_k$ is continuous but the inclusion of some contamination with the data spoils the normality assumption of error structure. One of these plots is the plot of the actual data, and the remaining are null plots, produced by simulating data from a null model that assumes $H_0$ is true. Which plot is the most different from the others, in the sense that there is the steepest slope?}
       \label{fig:test_contaminated}
\end{figure}

@@

%\begin{figure*}[hbtp]
%   \centering
%       \scalebox{0.6}{\includegraphics{power_observed.pdf}}
%       \caption{Observed power of visual test from Equation \ref{eqn:power_estimate} with pointwise 95\% confidence limits and the power of UMP test for sample size $n= 100,300$ and $\sigma = 12,5$ as per experiment 1.}
%       \label{fig:power_observed}
%\end{figure*}

@@

%\begin{figure*}[hbtp]
%   \centering
%       \scalebox{0.6}{\includegraphics{power_observed_exp3.pdf}}
%       \caption{Observed power of visual test from Equation \ref{eqn:power_estimate} with pointwise 95\% confidence limits and the power of UMP test for sample size $n= 100$ and $\sigma = 5$ as per experiment 3. }
%       \label{fig:power_observed_exp3}
%\end{figure*}

@@

\green{Would it make sense to expand this table now to show number of lineups for different experimental conditions?? Maybe not}
\begin{table}[hbtp]
\caption{Number of subjects, gender and total lineups for all three experimental data sets. Note that in some of the lines the number of male and female participants does not add up to the total number of participants due to missing demographic information. }
\begin{center}
\begin{tabular}{rrrrr|rrrr|rrrr}
 & \multicolumn{4}{c} {Experiment 1}  & \multicolumn{4}{c} {Experiment 2}  & \multicolumn{4}{c} {Experiment 3} \\ \cline{2-5}  \cline{6-9}   \cline{10-13}
 & Subj & Male & Fem & Total & Subj & Male & Fem & Total & Subj & Male & Fem & Total \\ 
   & 239 & 121 & 107 & 2317 & 351 & 185 & 164 & 3858 & 156 & 106 &  50 & 1681 \\ 
   \hline
\end{tabular}
\end{center}
\label{tbl:summary}
\end{table}

@@

@@

For each parameter combination, the effect of the lineup is computed as $$E=\beta*\sqrt {n}/\sigma.$$ The experimental data is displayed in figure \ref{fig:power_loess_effect} for different effect sizes. The loess curves fitted to the response data provide the estimated observed power for each test. The corresponding power curves of conventional tests are shown for comparison. The observed power for experiment 2 (scatter plot) is better than what we observed for experiment 1 (box plot). For experiment 3, observed power is always above the conventional test power since for contaminated data we can visually see a trend while conventional test fail to address that.

Figure \ref{fig:power_loess_effect} also reveals that at effect $E=0$, the power is close to 0.05 (Type-I error) for both experiment 1 and 2. For experiment 3 the scenario is different since contamination in the data is visible and this sets the lower limit of proportion correct at a higher value. Thus for fitting model \ref{mixed} to the data we set offset = $log(0.05/0.95)$ so that the estimated power has a fixed lower limit = 0.05 (Type-I error) for effect $E=0$. While fitting model \ref{mixed} we set $W_{\ell i} = E_{\ell i}$, the effect of lineup $\ell$ that subject $i$ has evaluated. Also $E$ is used as the only fixed effect covariate without intercept but offset for experiment 1 and 2. For experiment 3 we fit the intercept to allow the power curves to go above 0.05 as we see in figure \ref{fig:power_loess_effect}.

The power curves obtained by fitting model \ref{mixed} to the data for all the experiments applying different screening criteria is shown in figure \ref{fig:power_screening}. This shows how the screening criteria may affect the results. The corresponding power curves of conventional test are also shown. Notice that no mater what criteria we apply the result does not change much for experiment 2. Also, estimated power curve with criteria 5 for experiment 3 is above the conventional test power curve.

\begin{figure}[hbtp]
   \centering
       \scalebox{0.55}{\includegraphics{power_loess_effect.pdf}}
       \caption{Observed power shown by loess smoother with simultaneous bootstrap confidence band and the power of conventional test. Responses are 1 and 0 depending on the success or failure respectively to identify the actual plot in the lineup. Size of the dots represents the number of responses for each lineup. \blue{Response to the intuition of Di: We are overestimating the effect for experiment 3 here because to compute effect I used the beta values of the linear part of the contaminated data. True effect should be much smaller since overall contaminated data should have a smaller slope. Thus the effect of 5 that the plot is showing now would actually be effect 2 (may be). That means the  power that we see in the plot for effect 5 is actually the power for effect 2. How can we get true effect for this data?
}}
       \label{fig:power_loess_effect}
\end{figure}

%
%\begin{figure}[hbtp]
%   \centering
%       \scalebox{0.70}{\includegraphics{power_mixed.pdf}}
%       \caption{The overall power estimated from model \ref{mixed} are shown. The corresponding power curve for conventional test is shown for comparison.}
%       \label{fig:power_mixed}
%\end{figure}

%Each participant was shown a sequence of 10 lineup plots.  In total, 3629 lineups were evaluated by 324 people coming from many different locations across the globe.  The results of the experiment are summarized in Figure \ref{fig:power_observed} which shows the observed power from the survey data calculated using Equation \ref{eqn:power_estimate} along with 95\% confidence interval calculated using Fisher's exact method.

>>>>

\subsection{Subject-specific abilities}
Calculating the power  is intricately related to the probability to identify the data plot from the lineup.

@@

Let $X_{\ell i} \sim Binom_{1, p_{\ell i}}$ be the random variable, with $X_{ \ell i}= 1$, if  subject $i$ correctly identifies the data plot from lineup $\ell$, and 0 otherwise.
We can model $p_{ \ell i}$ with a logistic regression with subject-specific random effects as:
\begin{equation} \label{eqn:mixed}
g( p_{\ell i} )= W_{\ell i} \delta +  Z_{\ell i}  \tau_{\ell i},  
\end{equation}

@@

\begin{tabular}{lp{5.5in}}
$W$  & is a design matrix of covariates corresponding to specifics of lineup $\ell$ and subject $i$. Covariates could include  demographic information of individuals, such as age, gender, education level etc. as well lineup-specific elements, e.g. difficulty level. \\
$\delta$ & is a vector of parameters for fixed effects.\\
$Z_{\ell i}$ & $1 \le i \le K$, $1 \le \ell \le L$  is a design matrix corresponding to random effects specific to individual $i$ and lineup $\ell$.  \\
$\tau$ & is a vector of independent real-valued random variables $\tau_{\ell i}$,  $\tau  \sim  MVN(0,\sigma_\tau I_{KL \times KL})$. $\tau$ will usually include a component incorporating an individual's ability or skill to evaluate lineups.\\
\end{tabular}
}

The model in equation \eqref{eqn:mixed} provides the subject and lineup specific probability of successful evaluation as 
\begin{equation} \label{eqn:mixed_power}
\hat p_{\ell i} = \frac{exp(\hat g_{\ell i})}{1+ exp(\hat g_{\ell i})}
\end{equation}
where $\hat g_{\ell i} = W_{\ell i} \hat {\delta} +  Z_{\ell i}  \hat {\tau}_{\ell i}$.

%\red{This model is  described in more detail in equation \eqref{eqn:mixed}.}

@@

\noindent
\subsection{Lineup size, $m$}

One of the big differences between visual inference and the conventional testing situation is the use of a finite number $m-1$ of representatives of the null distribution as comparison against the test statistic. The choice of $m$ has an obvious impact on the test, and we will discuss this in the following paragraph.

@@

%Thus the density of $p_D$ is Uniform(0,1)  under $H_0$. As noted by \cite{Ruppert:2007}, under $H_1$ the density of $p_D$ $$f_{p_D}(p_D; \delta)= \frac{f_{|t|}(F_{|t|}^{-1}(1-p_D);\delta)}{f_{|t|}(F_{|t|}^{-1}(1-p_D))}$$ derived from equation \eqref{dist_p}  is a right skewed distribution.

@@

Figure \ref{fig:pval_power} gives an overview of the probability of picking the actual data plot for lineups of different size: as $m$ increases we have an increased probability to observe a more highly structured null plot by chance. We see that for a $p$-value $p_D$ of about 0.15 for the data plot the signal in the plot is so weak, that we cannot distinguish the actual data plot from null plots in a lineup of size $m=20$. % This pattern is confirmed by our findings described in Section~\ref{sec:simulation}.
%also seen in our experiments as shown in figure \ref{fig:pval_pcorrect}.

@@

Note that the \hh{left}  hand side in equation \eqref{eqn:power_estimate} is independent of the parameter $\theta$, but just based on the lineup evaluation by independent observers. While this allows us to compute this value independently of the parametric setting, it also means that observers might pick out the data plot for a reason not related to the value of $\theta$.

@@

%From model \ref{eqn:mixed} we obtain %$\pi=Pr(Y=1|X_1, X_2, ..., X_p)$, 
%the power of the underlying testing procedure as a population average %by ignoring the random effect part. 
%%Thus we have the overall probability of success ($\pi$ or power) 
%for specified sample size ($n$) and  variance ($\sigma$)  as 
%
%\begin{equation}\label{eqn:power} 
%Power(\beta) = \pi=Pr(Y=1|\beta, n, \sigma) 
%\end{equation}
%
%The main covariate factors that determine the power of the test are sample size ($n$), $\sigma$ and the parameter $\beta$. combining these we obtain one single covariate effect denoted by $E$ and defined as $E = \beta/ \sigma*\sqrt{n}$. Using equation \eqref{eqn:mixed}, we then have the power as
%
%\begin{equation}\label{eqn:effect} 
%Power(E) = \pi=Pr(Y=1| E) 
%\end{equation}

@@

% Di: This is specific for the regression model
%Consider the hypothesis test $H_0: \beta_k=0$ against $H_1: \beta_k \ne 0$ in model \ref{multi}.
%%Now consider estimating the power of the visual test. 
%%Suppose $F_{|t|}(.)$ is the distribution function of an absolute value of $t$. For the regression slope parameter $\beta$ suppose the associated test statistic in classical test be $t_{obs}$ with $p$-value $p_D$. By definition we have 
%%
%%$$p_D=Pr(|t| \ge t_{obs}| H_0)=1-F_{|t|}(t_{obs}) \Rightarrow |t_{obs}|=F_{|t|}^{-1}(1-p_D)$$
%%
%%\red{HH: the following paragraph should probably move up, but I'm not sure how to change it appropriately. Mahbub, could you take care of that?}
%%Now suppose $F_{|t|}(.;\delta)$ denotes the distribution function of an absolute value of $t$ distribution with non-centrality parameter $\delta$. Thus the distribution function of $p-value$ be  
%%\begin{eqnarray}\label{dist_p}
%%F{p_D}(p) &=& Pr(p_D \le p)=1-Pr(1-p_D \le 1-p) \nonumber \\
%%  &=& 1-Pr(F_{|t|}^{-1}(1-p_D) \le F_{|t|}^{-1}(1-p) ) \nonumber \\
%%  &=& 1-Pr(|t_{obs}| \le F_{|t|}^{-1}(1-p)) \nonumber \\
%%  &=&\left\{ \begin{array}{ll}
%%          1-F_{|t|}( F_{|t|}^{-1}(1-p))=p &\mbox{ ; under $H_0$} \\
%%          1-F_{|t|}( F_{|t|}^{-1}(1-p); \delta) &\mbox{ ; under $H_1$} 
%%       \end{array} \right.     
%%\end{eqnarray}
%%
%%Thus the density of $p_D$ is Uniform(0,1)  under $H_0$. As noted by \cite{Ruppert:2007}, under $H_1$ the density $$f_{p_D}(p_D; \delta)= \frac{f_{|t|}(F_{|t|}^{-1}(1-p_D);\delta)}{f_{|t|}(F_{|t|}^{-1}(1-p_D))}$$ derived from equation \eqref{dist_p}  is a right skewed distribution.  
%In a lineup plot we simulate $m-1$ residual data sets from null model where each of these $m-1$ null data sets produces a corresponding $p$-value $p_{0,i}$ and $p_{0,i} \sim \text{Uniform}(0,1)$ for $i = 1, ..., m-1$. Suppose  $p_0=\min_i(p_{0,i})$. Thus $p_0 \sim \text{Beta}(1,m-1)$.

@@

\begin{equation}\label{power_exp} 
   Power(\beta)=Pr(p_{D} < p_0)  \quad \text{for}  \quad \beta \ne 0
\end{equation}
Figure \ref{fig:power_expected} shows the power of the conventional test in comparison to the expected power of the visual test obtained from equation \eqref{power_exp}. 
Figure \ref{fig:power_expected} shows the power of the conventional test in comparison to the expected power of the visual test obtained from Equation \ref{power_exp}. Notice that the expected power of the visual test is %almost as good as 
\hh{close to} the power of the conventional test.

%Theorem \ref{thm:power} helps to compute the expected power of visual inference.
%\begin{thm}\label{thm:power}
% equation \eqref{power_exp}  yields expected power of visual inference as Power($\beta$) = $E(1-p_D)^{m-1}$ .
%\end{thm}
% 
%{ \em proof}
%
%Since $p_0 \sim \text{Beta}(1,m-1)$,  for $t \in (0,1)$ we have the distribution function of $p_0$ be
%\begin{eqnarray*}
%F{p_0}(t) &=& (m-1) \int_0^t{(1-p_0)^{(m-2)}dp_0} \\
%  &=&  -(m-1) \int_1^{1-t}{x^{(m-2)}dx}\\
%  &=& \left[ x^{m-1}\right]^1_{1-t}\\
%  &=&1-(1-t)^{m-1} \rightarrow 1 \quad \text{as} \quad m \rightarrow \infty
%\end{eqnarray*}
%
%% verified this computation from http://www.wolframalpha.com using command int (m-1)*(1-p)^(m-2) dp, p=0,t
%
%Thus expected power in equation \eqref{power_exp} be
%\begin{eqnarray*}
% Pr(p_{B} < p_0) &=& 1- Pr( p_0 \le p_{B} ) \\
%  &=& 1- \int_{0}^{1}{Pr( p_0 \le p_{B} |p_{B} =t) f_{p_{B} }(t)dt } \\
%  &=& 1-  \int_{0}^{1}{F{p_0}(t) f_{p_{B} }(t)dt } \\
%  &=& E(1-p_D)^{m-1}   \\
%%  & \rightarrow &1-  \int_{0}^{1}{ f_{p_{B} }(t)dt }  = F_{p_D}(0) \quad \text{as} \quad m \rightarrow \infty
%\end{eqnarray*}
%
%
\begin{figure}[hbtp]
   \centering
       \scalebox{.6}{\includegraphics{power_expected.pdf}}
       \caption{Expected power of visual test  and the power of conventional test for sample size $n =100$, $\sigma = 12$ and $m$=20. }
       \label{fig:power_expected}
\end{figure}

@@

\begin{figure}[p]
%\begin{figurehere}
   \centering
       \includegraphics[width=0.95\textwidth]{lineup_continuous.pdf}
       \caption{Lineup plot ($m=20$) using scatter plots for testing $H_0: \beta_k=0$ where covariate $X_k$ is continuous. One of these plots is the plot of the actual data, and the remaining are null plots, produced by simulating data from a null model that assumes $H_0$ is true. Which plot is the most different from the others, in the sense that there is the steepest slope?}
       \label{fig:test_continuous}
\end{figure}

@@

\[
  Y_i = \left\{
  \begin{array}{l l}
    \alpha+\beta X_i + \epsilon_i  & \quad  X_i \sim N(0,1) \quad  i =1,...,n\\
    \lambda+ \eta_i & \quad X_i \sim N(\mu,1/3) \quad  i=n+1, ...,n*c\\
  \end{array} \right.
\]
where $\epsilon_i \stackrel{iid}\sim N(0,\sigma)$, $\eta_i \stackrel{iid}\sim N(0,\sigma/3)$ and $\mu = -1.75$, $c$ is the proportion of contaminated data. For the experiment we consider $n=100$, $c=0.15$ producing actual data with $115$ points. Also $\alpha=0$ and $\lambda=10$, and $\sigma$ is chosen to be approximately 3.5, so that error standard deviation across both groups of the data is $5$. %Figure \ref{fig:cont_dat} shows an example of contaminated data, generated by this model for $\beta=5$.  
A linear model (Equation \ref{multi} with $p=1$  and intercept $\beta_0=0$) is fit to the contaminated data. This experiment also follows the test in Table \ref{tbl:stat_multiple}, case 2. The actual data plot shows a scatterplot of the residuals vs $X_1$, and the null data plots are scatterplots of null data generated by simulating residuals from $N(0, {\hat{\sigma}}^2$) and plotting against $X_1$.

@@

An example lineup for slope $\beta=0.4$ is shown in Figure \ref{fig:test_contaminated}.  Can you pick which plot is different? %Plot 7 is the actual data plot.

%\begin{figure}[hbt]
%\begin{figurehere}
%   \centering
%       \scalebox{0.6}{\includegraphics{contaminated_data.pdf}}
%       \caption{Scatterplot of a simulated contaminated data set generated such that a simple linear model produce a slope almost negligible, ie, if we fit model (Equation \ref{multi}) with $p=1$  to this data, the resulting estimate of $\beta_1$ would not be statistically significant. The conventional t-test $p$-value for the slope is 0.67 for this data set but visually we can see a convincing evidence of positive slope. 
%}
%       \label{fig:cont_dat}
%\end{figure}

\begin{figure}[p]
%\begin{figurehere}
   \centering
       \includegraphics[width=0.95\textwidth]{lineup_contaminated.pdf}
       \caption{Lineup plot ($m=20$) using scatter plots for testing $H_0: \beta_k=0$ where covariate $X_k$ is continuous but the inclusion of some contamination with the data spoils the normality assumption of error structure. One of these plots is the plot of the actual data, and the remaining are null plots, produced by simulating data from a null model that assumes $H_0$ is true. Which plot is the most different from the others, in the sense that there is the steepest slope?}
       \label{fig:test_contaminated}
\end{figure}

@@

%\begin{figure*}[hbtp]
%   \centering
%       \scalebox{0.6}{\includegraphics{power_observed.pdf}}
%       \caption{Observed power of visual test from equation \eqref{eqn:power_estimate} with pointwise 95\% confidence limits and the power of UMP test for sample size $n= 100,300$ and $\sigma = 12,5$ as per experiment 1.}
%       \label{fig:power_observed}
%\end{figure*}

@@

%\begin{figure*}[hbtp]
%   \centering
%       \scalebox{0.6}{\includegraphics{power_observed_exp3.pdf}}
%       \caption{Observed power of visual test from equation \eqref{eqn:power_estimate} with pointwise 95\% confidence limits and the power of UMP test for sample size $n= 100$ and $\sigma = 5$ as per experiment 3. }
%       \label{fig:power_observed_exp3}
%\end{figure*}

@@

\begin{table}[hbtp]
\caption{Number of subjects, gender and total lineups seen, and number of distinct lineups for all three experimental data sets. Note that in some of the lines the number of male and female participants does not add up to the total number of participants due to missing demographic information. }
\begin{center}
\begin{tabular}{ccrrcc}
   \hline 
 Experiment & Subject & Male & Female & Responses &Lineup\\ 
    \hline
  1 & 239 & 121 & 107 & 2317 & 60 \\ 
  2 & 351 & 185 & 164 & 3858 & 70 \\ 
  3 & 156 & 106 &  50 & 1681 & 30 \\ 
   \hline
\end{tabular}
\end{center}
\label{tbl:summary}
\end{table}

@@

\subsection{Model fitting}
For each parameter combination, the effect is computed as

\[
E=\beta (\sqrt {n}/\sigma).
\]

The model in equation \eqref{eqn:mixed} is fit using $E$ as the only fixed effect covariate without intercept. we set offset = $log(0.05/0.95)$ so that the estimated power has a fixed lower limit at 0.05 (Type-I error) when $E=0$. Different subjects are expected to have different skill levels in evaluating lineups. To account for subject specific variation we use subject as random effect. We also use an interaction between subject and effect $E$ to allow for subject specific random slope. This leads to setting  $W_{\ell i} = E_{\ell i}$, the effect of lineup $\ell$ that subject $i$ has evaluated in equation \eqref{eqn:mixed}.  While for experiment 1 and 2 we don't allow intercept but offset, for experiment 3 an intercept is estimated to allow the power curves to go above 0.05 when $E=0$. This is because forcing power to be fixed at 0.05 for $E=0$ is not justified by the experiment 3 data (figure \ref{fig:power_loess_effect}).

For all the three experiments, each subject saw a lineup just once. This doesn't allow fitting a lineup random effect, in addition to a subject random effect, but it does simplify the model since $\tau_{\ell i} = \tau_i$. The subject-specific probability ($p_i$) of successful evaluation of a lineup is then estimated using equation \eqref{eqn:mixed_power}. For computation we use R \citep{R} package lme4.

Table \ref{tbl:model_par} shows the parameter estimates of the mixed effects model of the subject-specific variation.

\green{This section needs to have sufficient detail of the model fitting, output and diagnostics to feed into the next few sections discussing the findings.}

\begin{table}[hbtp]
\caption{Parameter estimates of model \ref{eqn:mixed}.}
\begin{center}
\begin{tabular}{ccc}
  \hline
 &  \multicolumn{2}{c} {Estimates}  \\
 \cline{2-3}
 Experiment & Effect & inter subject correlation\\
  \hline
  1 & 0.39 &  \\ 
  2 & 1.22 &  \\ 
  3 &  & \\ 
   \hline
\end{tabular}
\end{center}
\label{tbl:model_par}
\end{table}

@@

\noindent Figure \ref{fig:power_loess_effect} shows the power against effect for the three experiments. Data values are represented as dots with size indicating count at this value. The loess curves fit the data, estimating the observed power for different effect sizes, with grey bands indicating simultaneous bootstrap confidence bands. For comparison, the dashed lines show the corresponding power curves of the conventional tests. The observed power for experiments 1 and 2 are lower than the power of the conventional test. This is not a surprise -- it is not expected that visual inference will beat conventional tests, in the scenario when assumptions are satisfied. It is encouraging to see that visual inference mirrors the power vs effect relationship of conventional testing, and comes quite close in experiment 2, where scatterplots are used as the test statistic.

\begin{figure}[hbtp]
   \centering
       \scalebox{0.55}{\includegraphics{power_loess_effect.pdf}}
       \caption{Power in comparison to effect for the three experiments. Points indicate subject responses, with size indicating count. Responses are 1 and 0 depending on the success or failure respectively to identify the actual plot in the lineup. The loess curve (blue) estimates the observed power, and grey bands show simultaneous bootstrap confidence band.  Conventional test power is drawn as a dashed line. For experiment 3, conventional power is based on the slopes of the linear part of the non-contaminated data. Power for contaminated data is shown by cross mark.
}
       \label{fig:power_loess_effect}
\end{figure}

Results for experiment 3, are quite different. This is the situation where we expect to see the potential of visual inference, and indeed we do: the power of visual inference is always high, and much higher than the conventional test at small effect sizes. There is no actual conventional power in this situation, because assumptions are violated. It is computed assuming that there is no contamination, using the non-contaminated part of the simulation model. The  conventional power of true effect of the contaminated data is computed anyway and presented in figure \ref{fig:power_loess_effect} by cross marks. For this we obtain the estimated slope for the contaminated data and compute the estimated effect for which conventional power is calculated. Notice that the power structure is completely broken for contaminated data. For smaller original effect we get huge power and that explains why we see so many small $p$-values for experiment 3

It is really curious to see in the results of experiment 3 that the power of the visual test is really high when effect is small. These results are based on whether the subject correctly picked the actual data plot, regardless of reason. Although they were asked to select the plot that exhibited the highest association between the two variables, it may have been that they were cueing to the cluster of contaminated data. This will be explored further in Section \ref{sec:TypeIII}.

%(scatter plot) is better than what we observed for experiment 1 (box plot). For experiment 3, observed power is always above the conventional test power since for contaminated data we can visually see a trend while conventional test fail to address that.

It can also be seen that at effect $E=0$, the power is close to 0.05 (Type-I error) for both experiments 1 and 2. For experiment 3 the scenario is different. The contamination in the data is visible which sets the lower limit of proportion correct at a higher value.

%The power curves obtained by fitting model \ref{eqn:mixed} to the data for all the experiments applying different screening criteria is shown in figure \ref{fig:power_screening}. This shows how the screening criteria may affect the results. The corresponding power curves of conventional test are also shown. Notice that no mater what criteria we apply the result does not change much for experiment 2. Also, estimated power curve with criteria 5 for experiment 3 is above the conventional test power curve.
